---
title: "Turtle KEGG analysis"
author: "Yuejiao"
date: "2025-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/hsf378/PhD/Analysis/genomic-novelty")
knitr::opts_chunk$set(echo = TRUE)
library(clusterProfiler)
library(tidyverse)
library(stringr)
library(ggplot2)
library(pathview)
library(svglite)
library(KEGGREST)
library(cowplot)
```

## Load the kegg db

```{r}
cmy_kegg_df <- data.frame(cmy_id = names(keggConv("ncbi-geneid", "cmy")),
          ncbi_id = keggConv("ncbi-geneid", "cmy")) %>% 
  separate(.,ncbi_id, into = c("info", "ncbi_gid"), sep = ":") %>% 
  select(-info) 
kegg_list <- list()
for (i in c("pss","ciz","cmy","ccay","dcc","cpic","tst","cabi","mrv")){
  i.kegg_df <- data.frame(id = names(keggConv("ncbi-geneid", i)),
          ncbi_id = keggConv("ncbi-geneid", i)) %>% 
  separate(.,ncbi_id, into = c("info", "ncbi_gid"), sep = ":") %>% 
  select(-info) 
  kegg_list[[i]] <- i.kegg_df
}

turtle_species <- c("pss","ciz","cmy","ccay","dcc","cpic","tst","cabi","mrv") 

all_kegg_links <- list()

for (sp in turtle_species) {
  message("Processing ", sp, "...")
  links <- keggLink("pathway", sp)  # returns gene → pathway mapping
  df <- data.frame(
    gene = gsub("^.*:", "", names(links)),
    pathway = gsub("^.*:", "", as.character(links))
  )
  all_kegg_links[[sp]] <- df
}
all_kegg_links[["mrv"]] %>% filter(gene == "120368533")
links
```

## Load background gene_list

```{r}
gene_list <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/candidate_enricher_gene_info.txt", 
                      col_names = c("gene","ncbi_gid","info")) %>% 
  select(-info) %>% 
  mutate( ncbi_gid = as.character(ncbi_gid))
gene_input <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/candidate_enricher_og2gene.txt",
                       col_names = c("busco_id","gene"))
genes <- left_join(gene_input,gene_list, by = "gene") %>% 
  left_join(.,cmy_kegg_df, by = "ncbi_gid" ) %>% 
  filter(!is.na(cmy_id ))
```

## KEGG enrichment

### load outiler genes
#### pair outiler genes to green sea turtle NCBI Gene IDs 
```{r}
### cmy as background
# 255 genes
outliers_dn <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dn_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,genes, by = "busco_id") %>% 
  filter(!is.na(cmy_id )) %>% 
  select(ncbi_gid,cmy_id)
# write_tsv(outliers_dn,"~/Downloads/trutle_project_outliers_dn.tsv")

# 255 genes, converted 248 genes , missing 108898at32523, 112465at32523, 168156at32523, 259072at32523, 283327at32523, 290126at32523, 291010at32523, while 48547at32523 has duplicate ncbi gene IDs. 
outliers_ds <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/ds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,genes, by = "busco_id") %>% 
  filter(!is.na(cmy_id )) %>% 
  select(ncbi_gid,cmy_id)
# write_tsv(outliers_ds,"~/Downloads/trutle_project_outliers_ds.tsv")

# 255 genes, converted 252 genes, missing 283327at32523, 290126at32523, 291010at32523, while 48547at32523 has duplicate ncbi gene IDs.
outliers_dnds <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dnds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,genes, by = "busco_id") %>% 
  filter(!is.na(cmy_id )) %>% 
  select(ncbi_gid,cmy_id)
# write_tsv(outliers_dnds,"~/Downloads/trutle_project_outliers_dnds.tsv")

# 255 genes, converted 254 genes, missing 291010at32523, while 48547at32523 has duplicate ncbi gene IDs.
outliers_pure <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/raw_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,genes, by = "busco_id") %>% 
  filter(!is.na(cmy_id )) %>% 
  select(ncbi_gid,cmy_id)
# write_tsv(outliers_pure,"~/Downloads/trutle_project_outliers_pure.tsv")

### Residual PC1, 255 genes, convert 251 genes, missing 108898at32523, 259072at32523,283327at32523,290126at32523
res_pc1 <- read.table(file = "Results/4_additional_plots/outliers_genes/pc1_resid_q95.csv", header = T) %>% 
  as.tibble(.) %>% 
  left_join(.,genes, by = c("x" = "busco_id")) %>% 
  filter(!is.na(cmy_id))%>% 
  select(ncbi_gid,cmy_id)

### Residual PC2, 255 genes, convert 250 genes, missing 108898at32523, 259072at32523, 277695at32523, 283327at32523, 290126at32523
res_pc2 <- read.table(file = "Results/4_additional_plots/outliers_genes/pc2_resid_q95.csv", header = T) %>% 
  as.tibble(.) %>% 
  left_join(.,genes, by = c("x" = "busco_id")) %>% 
  filter(!is.na(cmy_id)) %>% 
  select(ncbi_gid,cmy_id)

### Residual PC3, 255 genes, convert 251 genes, missing 108898at32523, 259072at32523, 283327at32523, 290126at32523
res_pc3 <- read.table(file = "Results/4_additional_plots/outliers_genes/pc3_resid_q95.csv", header = T) %>% 
  as.tibble(.) %>% 
  left_join(.,genes, by = c("x" = "busco_id")) %>% 
  filter(!is.na(cmy_id)) %>% 
  select(ncbi_gid,cmy_id)
```
#### pair outiler genes to Chinese soft-shelled turtle NCBI Gene IDs 
```{r}
pss_genes <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/pss_5031_merged.tsv", 
                      col_names = c("gene","busco_id","ncbi_gid")) %>% 
  mutate(pss_id = paste("pss",ncbi_gid,sep = ":"))
### pss as background
# 255 genes, converted 250 genes, missing 3576at32523, 30310at32523, 280970at32523, 246490at32523, 235309at32523.
outliers_dn_pss <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dn_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,pss_genes, by = "busco_id") %>% 
  arrange(desc(busco_id)) %>% 
  filter(!is.na(pss_id )) %>% 
  select(ncbi_gid,pss_id)
# write_tsv(outliers_dn,"~/Downloads/trutle_project_outliers_dn.tsv")

# 255 genes, converted 237 genes. 
outliers_ds_pss <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/ds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,pss_genes, by = "busco_id") %>% 
  arrange(desc(busco_id)) %>% 
  filter(!is.na(pss_id )) %>% 
  select(ncbi_gid,pss_id)
# write_tsv(outliers_ds,"~/Downloads/trutle_project_outliers_ds.tsv")

# 255 genes, converted 249 genes
outliers_dnds_pss <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dnds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,pss_genes, by = "busco_id") %>% 
  arrange(desc(busco_id)) %>% 
  filter(!is.na(pss_id )) %>% 
  select(ncbi_gid,pss_id)

# write_tsv(outliers_dnds,"~/Downloads/trutle_project_outliers_dnds.tsv")

# 255 genes, converted 248 genes
outliers_pure_pss <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/raw_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,pss_genes, by = "busco_id") %>% 
  arrange(desc(busco_id)) %>% 
  filter(!is.na(pss_id )) %>% 
  select(ncbi_gid,pss_id)
# write_tsv(outliers_pure,"~/Downloads/trutle_project_outliers_pure.tsv")
```

### enrich kegg function for different background
```{r}
enrich_ref_kegg <- function(df){
  out_df <- enrichKEGG(gene = unique(df$ncbi_gid),
         pAdjustMethod = "fdr",
         organism = "cmy",
         pvalueCutoff = 0.05)
  out_df@result$Description <- sub(" - Chelonia mydas \\(green sea turtle\\)", "", out_df@result$Description)
  return(out_df)
}

enrich_pss_kegg <- function(df){
  out_df <- enrichKEGG(gene = unique(df$ncbi_gid),
         pAdjustMethod = "fdr",
         organism = "pss",
         pvalueCutoff = 0.05)
  out_df@result$Description <- sub(" - Pelodiscus sinensis \\(Chinese soft-shelled turtle\\)", "", out_df@result$Description)
  return(out_df)
}
```

### KEGG analysis
#### dN results
```{r}
kegg_cmy_genes_dn <- enrich_ref_kegg(outliers_dn) ## 9 enriched terms
kegg_pss_genes_dn <- enrich_pss_kegg(outliers_dn_pss) ## 9 enriched terms
# Number of enriched terms
nrow(as.data.frame(kegg_ref_genes_dn))
nrow(as.data.frame(kegg_pss_genes_dn))
as.data.frame(kegg_pss_genes_dn)
as.data.frame(kegg_ref_genes_dn)
```
#### dS results
```{r}
kegg_ref_genes_ds <- enrich_ref_kegg(outliers_ds) #0
kegg_pss_genes_ds <- enrich_pss_kegg(outliers_ds_pss) #0
# Number of enriched terms
nrow(as.data.frame(kegg_ref_genes_ds))
```
#### dN/dS results
```{r}
kegg_ref_genes_dnds <- enrich_ref_kegg(outliers_dnds) #8 enriched terms
kegg_pss_genes_dnds <- enrich_pss_kegg(outliers_dnds_pss) #8 enriched terms
# Number of enriched terms
nrow(as.data.frame(kegg_ref_genes_dnds))
nrow(as.data.frame(kegg_pss_genes_dnds))


```

#### visualize the pathway and highligh the outlier genes in red for dn/ds results
```{r}
desired_order <-  c("Spliceosome","Basal transcription factors", "mRNA surveillance pathway", "Proteasome", "RNA degradation", "Polycomb repressive complex","Viral life cycle - HIV-1","Cell cycle")
kegg_ref_genes_dnds@result$Description <- factor(kegg_ref_genes_dnds@result$Description,levels = desired_order)
KEGG_enriched_pathways <- dotplot(kegg_ref_genes_dnds, 
        x = "GeneRatio", 
        showCategory = length(desired_order),
        color = "p.adjust") + 
  scale_y_discrete(limits = rev(desired_order)) +
  scale_color_gradientn(name = "Adjusted p-value", colors =c("#72315C", "#A6A867")) +
  #scale_size(name = "Gene count") +
  theme_minimal(base_family = "Arial") +
  labs(size="Gene count",
       x="Gene ratio") +
  theme(
    
    # base text (will cascade to anything not explicitly set below)
    text = element_text(
      family = "Arial", 
      color  = "black",
      size   = 14
    ),
    
    # axis titles and axis tick labels
    axis.title  = element_text(color = "black"),
    axis.text   = element_text(color = "black"),
    
    # legend title and labels
    legend.title = element_text(color = "black"),
    legend.text  = element_text(color = "black"),
    
    # strip text (for facets, if any)
    strip.text   = element_text(color = "black")
  )

## TODO: Arial

svglite('Plots/3_kegg_analysis/KEGG_enriched_pathways.svg', width = 10, height = 4)
print(KEGG_enriched_pathways)
dev.off()

KEGG_enriched_pathways
ggsave("Plots/3_kegg_analysis/KEGG_enriched_pathways.png", units = "in", height = 4, width = 6,
       dpi = 700)

#### write output enriched geneID
kegg_ref_genes_dnds@result %>% filter(p.adjust < 0.05) %>% select(geneID) %>% write_csv("Results/geneid.csv")

#### KEGG pathways 
kegg_path_id <- rownames(kegg_ref_genes_dnds@result %>% filter(p.adjust < 0.05) %>% select(ID))

for (i in kegg_path_id) {
  
  pathview(gene.data=outliers_dnds$ncbi_gid, 
         pathway.id=i, 
         species = "cmy")
}
```
#### visualize the pathway and highligh the outlier genes in red for dn/ds results using PSS as the background geneset
```{r}
kegg_pss_genes_dnds@result$Description <- factor(kegg_pss_genes_dnds@result$Description,levels = desired_order)
KEGG_enriched_pathways_pss <- dotplot(kegg_pss_genes_dnds, 
        x = "GeneRatio", 
        showCategory = length(desired_order),
        color = "p.adjust") + 
  scale_y_discrete(limits = rev(desired_order)) +
  scale_color_gradientn(name = "Adjusted p-value", colors =c("#72315C", "#A6A867")) +
  #scale_size(name = "Gene count") +
  theme_minimal(base_family = "Arial") +
  labs(size="Gene count",
       x="Gene ratio") +
  theme(
    
    # base text (will cascade to anything not explicitly set below)
    text = element_text(
      family = "Arial", 
      color  = "black",
      size   = 14
    ),
    
    # axis titles and axis tick labels
    axis.title  = element_text(color = "black"),
    axis.text   = element_text(color = "black"),
    
    # legend title and labels
    legend.title = element_text(color = "black"),
    legend.text  = element_text(color = "black"),
    
    # strip text (for facets, if any)
    strip.text   = element_text(color = "black")
  )
svglite('Plots/3_kegg_analysis/KEGG_enriched_pathways_pss.svg', width = 10, height = 4)
print(KEGG_enriched_pathways_pss)
dev.off()
```

#### pairwise distance results
```{r}
kegg_ref_genes_pure <- enrich_ref_kegg(outliers_pure) # 3
kegg_pss_genes_pure <- enrich_pss_kegg(outliers_pure_pss) # 3
```

#### Residual PCs results
```{r}
kegg_cmy_res_pc1 <- enrich_ref_kegg(res_pc1) #0
kegg_cmy_res_pc2 <- enrich_ref_kegg(res_pc2) #0
kegg_cmy_res_pc3 <- enrich_ref_kegg(res_pc3) #0
```


### multiple species KEGG enrichment

```{r}

# all KO pathways
pws <- keggList("pathway", "ko")          
# pathway→KO
links <- keggLink("ko", "pathway")       
# convert to dataframe
TERM2GENE <- data.frame(
  term = sub("path:", "", names(links)),
  gene = sub("ko:", "", links),
  stringsAsFactors = FALSE
) %>% filter(str_detect(term,"ko"))

TERM2NAME <- data.frame(
  term = sub("path:", "", names(pws)),
  name = as.vector(pws),
  stringsAsFactors = FALSE
)
```
##### KOs overlap between cmy and pss
```{r}
# 11339
length(unique(keggLink("ko", "cmy")))
# 10630
length(unique(keggLink("ko", "pss")))
# 10467
length(intersect(unique(keggLink("ko", "cmy")), unique(keggLink("ko", "pss"))))
```

## KofamKOALA result

```{r}
cpb_kegg <- read_tsv("Data/3_kegg_analysis/KOfamKOALA/Kofamscan results - 8478_all.tsv") ### 11171 unique KOs
### unique KOs
length(unique(cpb_kegg$KO))

tmt_kegg <- read_tsv("Data/3_kegg_analysis/KOfamKOALA/Kofamscan results - 1415176_all.tsv") ### 10471 unique KOs
### unique KOs
length(unique(tmt_kegg$KO))

cmy_kegg_kofamkoala <- read_tsv("Data/3_kegg_analysis/KOfamKOALA/Kofamscan results - 8469_all.tsv") ### 10514
### unique KOs
length(unique(cmy_kegg_kofamkoala$KO))

pss_kegg_kofamkoala <- read_tsv("Data/3_kegg_analysis/KOfamKOALA/Kofamscan results - 13735_all.tsv") ### 10089 KOs
### unique KOs
length(unique(pss_kegg_kofamkoala$KO))

#### KEGG database

cmy_kegg_r <- data.frame(KOfull=unique(keggLink("ko", "cmy"))) %>% 
  separate(.,KOfull, into=c("ko","KO"),sep=":") 
pss_kegg_r <- data.frame(KOfull=unique(keggLink("ko", "pss"))) %>% 
  separate(.,KOfull, into=c("ko","KO"),sep=":")

```

### Overlap between KofamKOALA result and KEGG database
```{r}
length(intersect(cmy_kegg_r$KO,cmy_kegg_kofamkoala$KO)) ### 10278
length(intersect(pss_kegg_r$KO,pss_kegg_kofamkoala$KO)) ### 9829
```

### Pool background KOs
```{r}
# union from KOfamKOALA results
pool_bg <- bind_rows(cpb_kegg,tmt_kegg, cmy_kegg_kofamkoala,pss_kegg_kofamkoala) %>%
  select(gene_name,KO)
## unique KOs
length(unique(pool_bg$KO)) ### 11,419 

# intersection
length(intersect(intersect(intersect(cpb_kegg$KO,tmt_kegg$KO),cmy_kegg_kofamkoala$KO),pss_kegg_kofamkoala$KO)) ## 9,140

# union from mixed KOfamKOALA and KEGG annotated
pool_bg_mix <- bind_rows(cpb_kegg %>% select(KO),
                         tmt_kegg %>% select(KO), 
                         cmy_kegg_r %>% select(KO),
                         pss_kegg_r %>% select(KO)) %>% 
  distinct(.) ### 11,770

```

## forward KOs
### load paired Busco gene and Ortho gene per species
```{r}
cpb_pairs <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/8478_0.candidate_og2gene.txt", 
                      col_names = c("busco_id","gene"))
tmt_pairs <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/1415176_0.candidate_og2gene.txt", 
                      col_names = c("busco_id","gene"))
cmy_pairs <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/8469_0.candidate_og2gene.txt", 
                      col_names = c("busco_id","gene"))
pss_pairs <- read_tsv("Data/3_kegg_analysis/KEGG og2geneid/13735_0.candidate_og2gene.txt", 
                      col_names = c("busco_id","gene"))
busco_pairs <- bind_rows(cpb_pairs,tmt_pairs,cmy_pairs,pss_pairs)
```
### pair outlier busco gene to species gene as forward gene set
```{r}
# 255 genes, converted 255 genes
outliers_dn_pairs <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dn_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,busco_pairs, by = "busco_id") %>% 
  left_join(.,pool_bg, by = c("gene" = "gene_name"))


# 255 genes, converted 254 genes , missing 259072at32523. 
outliers_ds_pairs <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/ds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,busco_pairs, by = "busco_id") %>% 
  left_join(.,pool_bg, by = c("gene" = "gene_name"))


# 255 genes, converted 255 genes
outliers_dnds_pairs <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/dnds_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,busco_pairs, by = "busco_id") %>% 
  left_join(.,pool_bg, by = c("gene" = "gene_name"))



# 255 genes, converted 255 genes
outliers_pure_pairs <- read.csv("Results/1_mahalanobis_outliers/outliers_genes/raw_q95.csv", col.names = c("busco_id","pval")) %>% 
  left_join(.,busco_pairs, by = "busco_id")  %>% 
  left_join(.,pool_bg, by = c("gene" = "gene_name"))


```

#### KEGG enrichment by multiple species as background
```{r}
dnds_kegg <- enricher(
  gene          = outliers_dnds_pairs %>% filter(!is.na(KO)) %>% pull(KO) %>% unique(),
  universe      = pool_bg %>% pull(KO) %>% unique(),
  TERM2GENE     = TERM2GENE,
  TERM2NAME     = TERM2NAME,
  pAdjustMethod = "fdr",
  pvalueCutoff = 0.05
)
as_data_frame(dnds_kegg)

```
```{r}
desired_order_poolbg <-  c("Spliceosome","Basal transcription factors", "mRNA surveillance pathway", "Proteasome", "RNA degradation", "Polycomb repressive complex","Viral life cycle - HIV-1","Cell cycle","Amyotrophic lateral sclerosis","Huntington disease","Spinocerebellar ataxia")
dnds_kegg@result$Description <- factor(dnds_kegg@result$Description,levels = desired_order_poolbg)
dnds_kegg_poolbg <- dotplot(dnds_kegg %>% filter(!is.na(Description)), 
        x = "GeneRatio", 
        showCategory = length(desired_order_poolbg),
        color = "p.adjust") + 
  scale_y_discrete(limits = rev(desired_order_poolbg)) +
  scale_color_gradientn(name = "Adjusted p-value", colors =c("#72315C", "#A6A867")) +
  #scale_size(name = "Gene count") +
  theme_minimal(base_family = "Arial") +
  labs(size="Gene count",
       x="Gene ratio") +
  theme(
    
    # base text (will cascade to anything not explicitly set below)
    text = element_text(
      family = "Arial", 
      color  = "black",
      size   = 14
    ),
    
    # axis titles and axis tick labels
    axis.title  = element_text(color = "black"),
    axis.text   = element_text(color = "black"),
    
    # legend title and labels
    legend.title = element_text(color = "black"),
    legend.text  = element_text(color = "black"),
    
    # strip text (for facets, if any)
    strip.text   = element_text(color = "black")
  )
combined_plot <- plot_grid(
  KEGG_enriched_pathways_pss,
  dnds_kegg_poolbg,
  nrow = 2,
  labels = c("A", "B")
)
svglite::svglite("Plots/3_kegg_analysis/KEGG_enriched_pathways_supplementFig11.svg",
                 width = 10, height = 8)
# gridExtra::grid.arrange(KEGG_enriched_pathways_pss,
#                         dnds_kegg_poolbg,
#                         nrow = 2)
print(combined_plot)
dev.off()


```

